---
title: "Classes"
author: "Yang Wu"
date: "10/5/2021"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.height = 6, fig.width = 7, fig.align = "center")
library(sloop)
library(magrittr)
```

## Classes

To make an object an instance of a class, we need to set the class attribute:

```{r}
# Create and assign class in one step
x <- structure(.Data = list(), class = "my_class")
# Object type
otype(x)
# Class
attributes(x)
```

Or, equivalently:

```{r}
# Create then set class
x <- list()
class(x) <- "my_class"
# Object type
otype(x)
# Class
attributes(x)
```

To determine the class of an S3 object and see if an object is an instance of a class:

```{r}
# Check class
class(x)
# Check if object is an instance of a class
inherits(x, what = "my_class", which = FALSE)
# The object is not of class "lm"
inherits(x, what = "lm", which = FALSE)
```

Three functions are important when creating a custom class:

* A constructor called `new_myclass()`

* A validator called `validate_myclass()`

* A helper called `myclass()`

---

# Constuctors

*The constructor creates new objects with a the correct structure for a given class*. This is because S3 does not have a formal definition of a class and so there is no built-in way to ensure that all objects of a given class have the same structure. The constructor should follow three principles:

* Be called `new_myclass()`

* Have one argument for the base object, and one for each attribute

* Check the type of the base object and the types of each attribute

For example, we can create a constructor for base classes--- the S3 class "Date." The "Date" class is built on double vectors with a single attribute--- its class "Date." 

```{r}
# One argument for the base object
new_Date <- function(x = double()) {
  # Check the type of the base object
  stopifnot(is.double(x))
  structure(x, class = "Date")
}
```

The value of the base object--- the double--- represents the number of days since `1970-01-01`, the Unix Epoch:

```{r}
new_Date(x = c(-1, 0, 7))
```

---

Another example is that for `difftime`, which is used to represent time differences. This object, also built on doubles, is of the class `difftime` and it has a `units` attribute.

```{r}
# Two arguments for base object and attributes, respectively
new_difftime <- function(x = double(), units = "secs") {
  # Check base type and type of attribute
  stopifnot(is.double(x))
  units <- match.arg(
    arg = units,
    choices = c("secs", "mins", "hours", "days", "weeks"),
    # Should arg be allowed to have more than one element?
    several.ok = FALSE
  )

  structure(.Data = x, class = "difftime", units = units)
}
```

Instantiate objects of class `difftime`:

```{r}
new_difftime(x = c(1, 20, 300), units = "secs")
new_difftime(x = c(300), units = "weeks")
```

A constructor will be used by experience users and so we can trade a bit of safety for performance. In other words, leave the more expensive checks for the validator.

---

# Validators

*The validator perform more computationally expensive checks, which may be required for more complicated classes*. For instance, the `factor` class require more complicated checks as the constructor, which only checks that the types are correct, may create malformed structures.

```{r}
# Constructor
new_factor <- function(x = integer(), levels = character()) {
  stopifnot(is.integer(x))
  stopifnot(is.character(levels))

  structure(.Data = x, class = "factor", levels = levels)
}
```

This example below will lead to malformed factors:

```{r, eval=FALSE}
# Error in as.character.factor(x) : malformed factor
new_factor(x = (1:8), "a")
```

Create a validator:

```{r}
validate_factor <- function(x) {
  # Get underlying base object
  # This would be an integer vector, each element representing a unique factor level
  values <- unclass(x)
  # Get levels attributes, which will be a character
  levels <- attr(x = x, which = "levels", exact = FALSE)

  # Check base object validity
  # The predicate function is.na(values) must return FALSE so that !is.na(values) returns TRUE
  # Both !is.na(values) AND values > 0 must be TRUE to avoid signalling the error condition object
  # If either !is.na(values) OR values > 0 is FALSE, all() will be FALSE and !all() will be TRUE
  # Use & rather than vectorized && since all elements of values must not be NA
  if (!all(!is.na(values) & values > 0)) {
    stop(
      "All 'x' values must be non-missing and greater than zero",
      call. = FALSE
    )
  }

  # Check attribute validity
  # The maximum of the values vector is the number of unique factor levels in the input x
  if (length(levels) < max(values)) {
    stop(
      "There must be at least as many 'levels' as possible values in 'x'",
      call. = FALSE
    )
  }

  # Return visibly the input
  return(x)
}
```

This validator can check for the validity of structure of the object created by the constructor:

```{r}
# The constructor creates a valid object of class factor
validate_factor(x = new_factor(x = 1:5, levels = letters[1:5]))
```

Or malformed factors:

```{r, eval=FALSE}
# Error: There must be at least as many 'levels' as possible values in 'x'
validate_factor(x = new_factor(x = 1:5, levels = c("male", "female")))
# Error: All 'x' values must be non-missing and greater than zero
validate_factor(x = new_factor(x = c(1:5, NA_integer_), levels = letters[1:6]))
# Error: All 'x' values must be non-missing and greater than zero
validate_factor(x = new_factor(x = 0:5, levels = letters[1:6]))
```

---

# Helpers

*A helper method provides a convenient way for others to create objects of a custom class*. A helper method should always have the following properties:

* Have the same name as the class, e.g., `myclass()`

* Finish by calling the constructor, and the validator, if it exists

* Create carefully crafted user interface with carefully chosen default values and useful conversions

Lastly, there are three common patterns:

* Sometimes the helper needs to coerce its inputs to the desired type. For instance, the constructor for objects of `difftime` class, `new_difftime()`, is very strict; it violates the usual convention that we can use an integer vector wherever we can use a double vector:

```{r, eval=FALSE}
# Error in new_difftime(x = 1:19, units = "hours") : is.double(x) is not TRUE
new_difftime(x = 1:19, units = "hours")
```

The constructor must be strict, but we may create a helper that coerces the input to a double:

```{r}
# Helper
difftime <- function(x = double(), units = "secs") {
  x <- as.double(x)
  # Finish by calling the constructor
  new_difftime(x, units = units)
}
# Now the integer input can be coerced
difftime(x = 1:19, units = "days")
```

---

* It may be convenient sometimes to specify a factor with a character vector. For objects of class `factor`, it may be helpful to create a helper method called `factor()` that takes a character vector and assumes that the levels should be the unique values. This is a simple implementation of the base function `factor`:

```{r}
factor <- function(x = character(), levels = unique(x)) {
  # Return the positions of (first) matches of x in table
  index <- match(x = x, table = levels)
  # Finish by calling constructor and validator
  validate_factor(x = new_factor(x = index, levels = levels))
}
```

We can use this helper to create objects of class `factor` by supplying character vectors:

```{r}
factor(x = c("male", "female", "other", "female", "male", "male"))
```

---

* Some complex objects are most naturally specified by multiple simple components. For instance, objects of classes `POSIXct` (built-on double vectors) and `POSIXlt`, i.e., date-time classes, may be constructed by supplying the individual components (year, month, day, etc.). We can create a `POSIXct()` helper method that resembles the `ISODatetime()` function or the `lubridate::make_datetime()` function:

```{r}
POSIXct <- function(year = integer(),
                    month = integer(),
                    day = integer(),
                    hour = 0L,
                    minute = 0L,
                    second = 0,
                    tzones = "") {
  lubridate::make_datetime(
    year = year,
    month = month,
    day = day,
    hour = hour,
    min = minute,
    sec = second,
    tz = tzones
  )
}
```

We can use this helper method to create objects of the date-time classes:

```{r}
# Create object
POSIXct(2020, 1, 1, tzone = "America/New_York")
# Base type
POSIXct(2020, 1, 1, tzone = "America/New_York") %>% typeof()
# Object type
POSIXct(2020, 1, 1, tzone = "America/New_York") %>% otype()
# Class
POSIXct(2020, 1, 1, tzone = "America/New_York") %>% attributes()
```

---

# Exercises 

## Exercise 1

Write a constructor for `data.frame` objects. What base type is a data frame built on? What attributes does it use? What are the restrictions placed on the individual elements? What about the names?

```{r}
# Base type
typeof(data.frame())
# Is OO object?
is.object(data.frame())
# Object type
otype(data.frame())
# Attributes
attributes(x = data.frame())
```

Create a constructor:

```{r}
new_data_frame <- function(x = list(), names = character(), row.names = integer()) {
  # Check base type
  stopifnot(is.list(x))
  # Check attribute types
  stopifnot(is.character(names))
  stopifnot(is.integer(row.names))

  structure(
    .Data = x,
    class = "data.frame",
    names = names,
    row.names = row.names
  )
}
```

We can use the constructor to create objects of class `data.frame`:

```{r}
new_data_frame(x = list(9:19, 32:42), names = c("x", "y"), row.names = 1:10)
```

---

## Exercise 2

Enhance the `factor()` helper to have better behavior when one or more values is not found in levels. What does `base::factor()` do in this situation?

The function `base::factor()` handles the situation where one or more values is not found in levels by coercing them to NA's:

```{r}
base::factor(sample(letters[1:5], size = 12, replace = TRUE), levels = letters[1:3])
```

To enhance our `factor()` helper method, we can use the `nomatch` argument in the `match()` function to capture the values that do not appear in the levels. Then use the captured values to signal a custom error condition object: 

```{r}
# Create custom abort function
abort_unknown_level <- function(x, levels = NULL, index) {

  # Capture values that did not have matches in levels
  unknown_levels <- x[!index] %>% unique()
  # Capture values that did have matches in levels
  matched_levels <- x[index] %>% unique()
  # Error message for user
  msg <- glue::glue(
    "Values {toString(unknown_levels)} are not found in levels; all unique values in 'x' must be found in ({toString(levels)})."
  )

  # Store meta data in condition call
  # This allows us to use rlang::catch_cnd() to get these meta data
  rlang::abort(
    "error_unknown_level",
    message = msg,
    unknown_levels = unknown_levels,
    matched_levels = matched_levels
  )
}
```

Now we can use the custom `abort()` function to throw the error in the `factor()` helper method:

```{r}
factor <- function(x = character(), levels = unique(x)) {

  # Return the positions of (first) matches of x in table
  # If any value in x has no match in table, then return FALSE
  # The index logical_index has TRUE's for the matched and FALSE's for the unmatched
  logical_index <- match(x = x, table = levels, nomatch = FALSE) %>% as.logical()

  # Custom condition object
  # If any element in logical_index is FALSE, all() will returns FALSE and !all() will return TRUE
  if (!all(logical_index)) {
    abort_unknown_level(x, levels, logical_index)
  }

  # If no error, finish by calling constructor and validator
  validate_factor(x = new_factor(x = index, levels = levels))
}
```

Let us see it in action. The example below will throw an error and we will use `rlang::catch_cnd()` to capture the error condition object:

```{r, eval=FALSE}
# This will return => Error: error_unknown_level
factor(
  x = sample(letters[1:5], size = 12, replace = TRUE),
  levels = letters[1:3]
)
```

We can examine the error condition object:

```{r}
# Capture condition object
error <- factor(
  x = sample(letters[1:5], size = 12, replace = TRUE),
  levels = letters[1:3]
) %>%
  rlang::catch_cnd()
# Examine message
error$message
# Examine metadata
error$unknown_levels
error$matched_levels
```

This structured condition object allows for more comprehensive user interface and stores more meta data for the developer.

---

## Exercise 3

Carefully read the source code of `factor()`. What does it do that our constructor does not?

This is source code our constructor:

```{r}
new_factor
```

This is source code of the `base::factor()` function: 

```{r}
base::factor
```

---

## Exercise 4

Factors have an optional “contrasts” attribute. Read the help for C(), and briefly describe the purpose of the attribute. What type should it have? Rewrite the new_factor() constructor to include this attribute.

```{r}
new_factor <- function(x = integer(),
                       levels = character(),
                       contrasts = NULL) {
  stopifnot(is.integer(x))
  stopifnot(is.character(levels))

  if (!is.null(constrasts)) {
    stopifnot(is.matrix(contrasts) && is.numeric(contrasts))
  }

  structure(
    x,
    levels = levels,
    class = "factor",
    contrasts = contrasts
  )
}
```

---

## Exercise 5

Read the documentation for `utils::as.roman()`. How would you write a constructor for this class? Does it need a validator? What might a helper do?

The function `utils::as.roman()` creates objects of class `roman`, which are internally built on integer vectors. This class has suitable methods for printing, formatting, sub-setting, coercion, etc.  

```{r}
# Object
utils::as.roman(x = 1:5)
# Base type
typeof(utils::as.roman(x = 1:5))
# Is OO object?
is.object(utils::as.roman(x = 1:5))
# Object type
otype(utils::as.roman(x = 1:5))
# Attributes
attributes(x = utils::as.roman(x = 1:5))
```

This object has one attribute--- its class `roman`. We can create a constructor for objects of this class as follows:

```{r}
new_roman <- function(x = integer()) {
  stopifnot(is.integer(x))
  structure(
    .Data = x, 
    class = "roman"
  )
}
```

We can use this constructor to create objects of class `roman`:

```{r}
new_roman(x = sample(x = 1:10, size = 5))
```

Since this is a simple class, we do not need a validator. However, since `double` and `integer` are used almost interchangeably by most R users, we may need to create a helper method that coerces double vector inputs to the desire type:

```{r}
roman <- function(x = integer()) {
  x <- as.integer(x)
  new_roman(x)
}
```

Now, if we supply doubles to the constructor, it would throw an error:

```{r, eval=FALSE}
# Error in new_roman(x = c(2.3, 423.2, 232.1)) : is.integer(x) is not TRUE
new_roman(x = c(2.3, 423.2, 232.1))
```

The helper method provides more flexibility:

```{r}
roman(x = c(2.3, 423.2, 232.1))
```

```{r, echo=FALSE}
rlang::env_unbind(env = rlang::caller_env(), nms = names(rlang::global_env()))
```













