---
title: "Generics and Methods"
author: "Yang Wu"
date: "10/6/2021"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = TRUE)
knitr::opts_chunk$set(warning = TRUE)
knitr::opts_chunk$set(fig.height = 6, fig.width = 7, fig.align = "center")
library(sloop)
library(tidyverse)
```

## Introduction

The S3 generic function performs **method dispatch**, that is, finding the specific implementation for a given class. Method dispatch is performed by `UseMethod()`, which is a primitive function. Every generic with the exception of internal (primitive) generics calls `UseMethod()` when performing method dispatch. The function has two arguments:

* "generic" is a character string naming a function (and not a built-in operator) and is required

* "object" is an object whose class will determine the method to be dispatched and this defaults to the first argument if it is not supplied

When a function calling `UseMethod("fun")` is applied to an object with class attribute `c("first", "second")`, the system searches for a function called `fun.first` and, if it finds it, applies it to the object. If no such function is found a function called `fun.second` is tried. If no class name in `c("first", "second")` produces a suitable function, the function `fun.default` is used, if it exists, or an error results.

```{r}
print
```

When `print()` is called on an object, for instance `matrix(1:20, 2)`, the class attribute `x("matrix" "array")`, the base type `c("integer")`, and the mode `c("numeric")` for this base object is used for method dispatch. Since neither of those functions `print.matrix`, `print.integer`, and `print.numeric` are found, `print.default` is used:

```{r}
s3_dispatch(call = print(matrix(1:20, 2)))
```

---

# Method dispatch

Essentially, `UseMethod()` creates a vector of method names:

```{r}
# Classes of x
x <- 1L
s3_class(x)
# Create vector of method names
paste0("generic", ".", c(s3_class(x), "default"))
```

Another example:

```{r}
# Object
x <- Sys.Date()
# Class
s3_class(x)
# Method dispatch
s3_dispatch(call = print(x))
```

* `=>` indicates the method that is called, here `print.Date()`
* `*` indicates a method that is defined, but not called, here `print.default()`

The “default” class is a special **pseudo-class**. This is not a real class, but is included to make it possible to define a standard fallback that is found whenever a class-specific method is not available.

---

# Finding methods

Find some base package S3 generic functions:

```{r}
# Base R functions
base_fn <- mget(x = ls(name = "package:base", all.names = TRUE), inherits = TRUE)
# Define a function to test whether a function is a S3 generic
test_fn <- function(x) {
  identical(ftype(x), c("S3", "generic"))
}
test_fn_safe <- safely(test_fn)
# Store results
results <- purrr::map(.x = base_fn, .f = test_fn_safe) %>% purrr::transpose()
# Create index
index <- purrr::map_lgl(.x = results[["result"]], .f = isTRUE)
# Subset to select S3 generic functions
base_fn %>%
  names() %>%
  `[`(index)
```

We can use `s3_methods_generic(x)` and `s3_methods_class(x)` where x is the name of class or generic.

```{r}
# Check all methods defined for a given generic
s3_methods_generic("mean")
s3_methods_generic("solve")
s3_methods_generic("within")
# Check all generics that have been provided methods for a given class
s3_methods_class("Date")
s3_methods_class("roman")
s3_methods_class("factor")
```

---

# Exercises

## Exercise 1

What generics does the `table` class have methods for?

```{r}
s3_methods_class(x = "table")
```

As an example, when an object of class `class` is passed to the generic function `plot()`, `plot.table()` is used:

```{r}
s3_dispatch(plot(table(sample(1:10, 20, replace = TRUE))))
```

---

## Exericse 2

What generics does the `ecdf` class have methods for?

```{r}
s3_methods_class(x = "ecdf")
```

---

## Exercise 3

Which base generic has the greatest number of defined methods?

```{r}
# List all names in base environment
# Character vector
ls(all.names = TRUE, env = base::baseenv()) %>%
  # Search for list of names in the base environment
  # List object
  mget(x = ., envir = baseenv(), inherits = FALSE) %>%
  # Keep elements when purrr::is_function() returns TRUE
  # List object
  keep(.x = ., .p = is_function) %>%
  # Get the names of the functions
  # Character vector
  names() %>%
  # Keep names that return TRUE when passed to is_s3_generic()
  # Character vector
  keep(is_s3_generic) %>%
  # The function s3_methods_generic(x) returns a tibble for each function name
  # The anonymous function returns a named integer vector each iteration of map
  # The two components of these vectors are the function names and the number of rows in the tibble
  # List object
  map(.x = ., .f = ~ set_names(x = nrow(s3_methods_generic(.x)), nm = .x)) %>%
  # Flatten the list object from vertical to horizontal
  flatten_int(.x = .) %>%
  # Sort
  sort(decreasing = TRUE) %>%
  head(n = 5)
```

---

## Exercise 4

Carefully read the documentation for `UseMethod()` and explain why the following code returns the results that it does. What two usual rules of function evaluation does `UseMethod()` violate?

```{r}
# Define g
g <- function(x) {
  x <- 10
  y <- 10
  UseMethod("g")
}
# Define method
g.default <- function(x) c(x = x, y = y)
# Bindings in global environment
x <- 1
y <- 5
# Method dispatch
s3_dispatch(call = g(x))
# Call g
g(x)
```

The rule that `UseMethod()` breaks is name-masking. Normally, names defined inside the function should mask names defined outside. In other words, the bindings `x <- 10` and ` y <- 10` inside `g()` should mask those defined outside--- `x <- 1` and ` y <- 1`. However, `g(x)` returns `c(1, 10)`, which means that the value of `x` is scoped from the global environment while that of `y` is scoped from within the generic function `g()`. Why?

* When `UseMethod("g")` is called within `g()`, method dispatch is performed and `g.default()` is called. Since `g.default()` is defined in the global environment, it binds or encloses the global environment; this is where `g.default()` scopes for the value of its argument `x`. Thus, the binding `x <- 1` is used. The argument supplied to the function `g()` is passed on as is and cannot be affected by code inside the generic. Since the binding `x <- 1` already exists in the enclosing environment of `g.default()`, `x <- 10` defined within the generic is masked.

---

## Exercise 5

What are the arguments to `[`? Why is this a hard question to answer?

```{r}
# Function type
ftype(`[`)
# Source code
pryr::show_c_source(.Primitive("["))
# Get methods for this generic
s3_methods_generic("[") %>%
  # Keep only those methods that are exported
  filter(.data = ., visible == TRUE)
```

---

# Object styles

Vector style classes have the key property that `length(x)` represents the number of observations in the vector. Some other variants do not have this property:

* Record style objects use a list of equal-length vectors to represent individual components of the object.The best example of this is POSIXlt, which underneath the hood is a list of 11 date-time components like year, month, and day. Record style classes override length() and subsetting methods to conceal this implementation detail.

```{r}
# Object
x <- as.POSIXlt(ISOdatetime(2020, 1, 1, 0, 0, 1:3))
x
# Length
length(x)
# Unclass results in a list of 11 elements, each of which is a date-time component
unclass(x)
# First date-time object
x[[1]]
# Class
x[[1]] %>% class()
# This object is a list under the hood
x[[1]] %>% unclass()
```

---

* Data frames are similar to record style objects in that both use lists of equal length vectors. 

```{r}
# Object
x <- data.frame(x = 1:100, y = 1:100)
# Length
length(x)
# Unclass
x %>% unclass()
# length
length(x %>% unclass() %>% `[[`(1))
```

---

* Scalar objects typically use a list to represent a single thing. For example, an lm object is a list of length 12 but it represents one model.

```{r}
# Object
mod <- lm(mpg ~ wt, data = mtcars)
# Length
length(mod)
```

---

# Exercises

## Exercise 1

Categorize the objects returned by `lm()`, `factor()`, `table()`, `as.Date()`, `as.POSIXct()`,`ecdf()`, `ordered()`, `I()` into the styles described above.

* Objects of class `lm` are scalar style

```{r}
# Object
mod <- lm(mpg ~ wt, data = mtcars)
# Class
class(mod)
# Base type
mod %>% typeof()
```

* Objects of class `factor` are vector style objects

```{r}
# Object
x <- factor(x = letters[1:10])
# Class
class(x)
# Unclass
unclass(x)
```

* Objects of class `table` are vector style objects

```{r}
# Object
x <- table(x = sample(x = 1:20, size = 50, replace = TRUE))
x
# Class
class(x)
# Unclass
unclass(x)
```

* Objects of class `Date` are vector style objects

```{r}
# Object
x <- as.Date(x = "2021-10-12")
x
# Class
class(x)
# Unclass
unclass(x)
```

* Date time objects of classes `POSIXlt` and `POSIXct` are record style objects

```{r}
# Object
x <- as.POSIXlt(Sys.time(), "America/New_York")
x
# Class
class(x)
# Unclass
unclass(x)
```

* Objects of class `ecdf` are cannot be categorized by any of the three categories--- scalar, record style, and vector style

```{r}
# Object
x <- rnorm(12) %>% ecdf()
x
# Class
x %>% attributes()
# Base type
# First class function object
typeof(x)
# Unclass
unclass(x) %>% attributes()
```

* Objects of classes c(`ordered`, `factor`) are vector style objects

```{r}
# Object
x <- ordered(letters[1:10])
# Class
x %>% attributes()
# Unclass
unclass(x)
```

* Objects of classes `AsIs` can be any style depending on the argument to `I(x)` 

```{r}
# Object
x <- I(mtcars)
y <- I(as.Date(x = "2021-10-12"))
# Class
attributes(x)
attributes(y)
```

---

## Exercise 2

What would a constructor function for `lm` objects, `new_lm()`, look like? Use ?lm and experimentation to figure out the required fields and their types.

```{r}
# Constructor for objects of class lm
new_lm <- function(x = list(), names = list()) {
  # Check base type and attributes
  stopifnot(is.list(x))
  stopifnot(is.list(names))

  structure(
    .Data = x,
    class = "lm",
    names = names
  )
}
```

The constructor is for developers and so experienced developers would know what fields are required in the `names` list. We may create a validator to ensure that the structure of the object created by `new_lm()` is as desired:

```{r}
# Validator for objects of class lm
validator_lm <- function(x) {

  # Validate each component in lm list
  if (is.double(x$coefficient)) {
    stop(
      "Coefficients must be of type 'double'"
    )
  }
  if (is.double(x$residuals)) {
    stop(
      "Residuals must be of type 'double'"
    )
  }
  if (is.double(x$effects)) {
    stop(
      "Effects must be of type 'double'"
    )
  }
  if (is.integer(x$rank)) {
    stop(
      "Coefficients must be of type 'integer'"
    )
  }
  if (is.double(x$fitted.values)) {
    stop(
      "Fitted.values must be of type 'double'"
    )
  }
  if (is.integer(x$assign)) {
    stop(
      "Assign must be of type 'integer'"
    )
  }
  if (is.list(x$qr)) {
    stop(
      "QR decomposition must be of type 'list'"
    )
  }
  if (is.integer(x$df.residual)) {
    stop(
      "Df.residual must be of type 'integer'"
    )
  }
  if (is.list(x$xlevels)) {
    stop(
      "Xlevels must be of type 'integer'"
    )
  }
  if (is.language(x$call)) {
    stop(
      "Call must be of type 'language'"
    )
  }
  if (purrr::is_formula(x$terms)) {
    stop(
      "Terms must be of type 'formula'"
    )
  }
  if (is.list(x$model)) {
    stop(
      "Model must be of type 'list'"
    )
  }

  x
}
```

We may also create a helper for the user, although the use of this is minimal, since we usually create objects of class `lm` from the actual data:

```{r}
# Helper
lm <- function(coefficients = double(),
               residuals = double(),
               effects = double(),
               rank = integer(),
               fitted.values = double(),
               assign = integer(),
               qr = list(),
               df.residual = integer(),
               xlevels = list(),
               call = lm(formula = y ~ x, data = data.frame()),
               terms = y ~ x,
               model = list()) {

  # Call constructor
  new_lm(
    x = list(),
    names = list(
      coefficients = coefficients,
      residuals = residuals,
      effects = effects,
      rank = rank,
      fitted.values = fitted.values,
      assign = assign,
      qr = qr,
      df.residual = df.residual,
      xlevels = xlevels,
      call = call,
      terms = terms,
      model = model
    )
  ) %>%
    # Call validator
    validator_lm(x = .)
}
```

---

Another solution:

```{r}
new_lm <- function(coefficients, residuals, effects, rank, fitted.values, assign,
                   qr, df.residual, xlevels, call, terms, model) {
  stopifnot(
    is.double(coefficients), is.double(residuals),
    is.double(effects), is.integer(rank), is.double(fitted.values),
    is.integer(assign), is.list(qr), is.integer(df.residual),
    is.list(xlevels), is.language(call), is.language(terms),
    is.list(model)
  )

  structure(
    list(
      coefficients = coefficients,
      residuals = residuals,
      effects = effects,
      rank = rank,
      fitted.values = fitted.values,
      assign = assign,
      qr = qr,
      df.residual = df.residual,
      xlevels = xlevels,
      call = call,
      terms = terms,
      model = model
    ),
    class = "lm"
  )
}
```
