---
title: "Purrr style"
author: "Ken Wu"
date: "7/26/2021"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(collapse = TRUE)
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.align = "center")
library(tidyverse)
options(digits = 7)
```

## Purrr style

1. Split `mtcars` data set into groups defined by the number of cylinders, using `split(x, f, drop = FALSE, ...)`, which divides the data in x (a vector or data frame) into subsets defined by f (factor). The argument drop accepts a logical vector of length 1 indicating if levels that do no occur in the factor should be dropped.

```{r}
by_cyl <- split(x = mtcars, f = as_factor(x = mtcars[["cyl"]]), drop = FALSE)
by_cyl[[2]]
```

Another way to subset the data is using `tidyr::nest(.data, ..., .names_sep = NULL)`, which creates a list-column of data frames called nested data-frames.

```{r}
by_cyl_tidyr <- mtcars %>%
  group_by(cyl) %>%
  nest()
glimpse(by_cyl_tidyr)
```

2. This creates a list of three data frames: cars with 4, 6, and 8 cylinders. The following code regresses mpg on weight using each subset of data, extracts the estimator of the parameter on weight:

```{r}
# Could use .x or . in the anonymous function
by_cyl %>%
  map(.f = ~ lm(mpg ~ wt, data = .x)) %>%
  # Pluck by name and position
  # We can do this since the the map() above returns a list
  # We pluck the elements named "coefficients" and select the second elements of those vectors
  map_dbl(.f = list("coefficients", 2))
```

---

## Base R style

* With pipe operator

```{r}
by_cyl %>%
  lapply(FUN = function(subset) lm(mpg ~ wt, data = subset)) %>%
  lapply(FUN = "coefficients") %>%
  vapply(FUN = function(x) x[[2]], FUN.VALUE = double(1))
```

* Without pipe operator

```{r}
# Use intermediate variables
list_of_models <- lapply(
  X = by_cyl,
  FUN = function(subset) lm(mpg ~ wt, data = subset)
)
# Extract estimators
vapply(
  X = list_of_models,
  FUN = function(model) model[["coefficients"]][[2]],
  FUN.VALUE = double(1)
)
```

---

## for loop

```{r}
# Pre-allocate container
estimators <- vector(mode = "double", length = length(x = by_cyl))
# For loop
for (i in seq_along(along.with = by_cyl)) {
  model <- lm(mpg ~ wt, data = by_cyl[[i]])
  estimators[[i]] <- pluck(.x = model, "coefficients", 2)
}
# Output
estimators
```
