---
title: "Handling conditions"
author: "Yang Wu"
date: "9/17/2021"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: false
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.align = "center")
library(rlang)
library(tidyverse)
options(digits = 4)
```

# Condition objects

The condition object can be captured using `rlang::catch_cnd()`:

```{r}
cnd <- rlang::catch_cnd(expr = stop(message = "An Error Please Stop"), classes = "condition")
str(cnd)
```

Built-in condition objects are *lists* with two elements:

* `message`, which is a length-1 character vector containing the text to display to a user. Use `base::conditionMessage()` to extract the message:

```{r}
base::conditionMessage(c = cnd)
```

*  `call`, which is the call which triggered the condition. We usually set `call. = FALSE` and so this will often be NULL. Use `base::conditionCall()` to extract it:

```{r}
new_cnd <- rlang::catch_cnd(expr = sum("e"), classes = "condition")
base::conditionCall(c = new_cnd)
```

Built-in conditions objects are usually those triggered when a built-in function throws an error. Condition objects have a `class` attribute, which makes them `S3` objects. The `class` attribute is a character vector, and it determines which handlers will match the condition:

```{r}
class(new_cnd)
```

---

## Exiting handlers

The function `tryCatch(expr, ..., finally)` registers exiting handlers, which allows us to override the default error behavior.

```{r}
# Function
f1 <- function(x) {
  base::tryCatch(
    # If the input "x" is invalid, a error condition object is created
    # Then, the control is passed to the handler--- the function below
    # The handler function runs and the the character vector "Error" is returned
    error = function(error) {
      "Error"
    },
    # If no error condition is signaled, run the code normally
    log(x, base = exp(1))
  )
}
# Test
# No condition
f1(3)
# Condition signaled
f1("3")
```

In addition, if the class of the condition object does not match the handler's name--- `error`, `message`, `warning`--- then the code also executes normally:

```{r}
# Matched condition
base::tryCatch(
  error = function(error) {
    "This is an error"
  },
  {
    base::stop("Error")
    10 + 7
  }
)
# Not matched
base::tryCatch(
  error = function(error) {
    "This is an error"
  },
  {
    base::message("Error")
    7 * sample(x = 1:20, size = 5, replace = TRUE)
  }
)
```

The nuance here is that after the condition is signaled, control is passed to the the context where `tryCatch()` was called. In other words, the code that throws the error and creates the error condition object is never run:

```{r}
tryCatch(
  message = function(message) {
    "Message"
  },
  {
    base::message("condition signaled")
    base::stop("This code is never run", call. = FALSE)
  }
)
```

The protected code `base::stop("This code is never run", call. = FALSE)` is run in the environment of `tryCatch()`, but the handler code `"Message"` is not since the handler `message` is a function itself.

```{r}
tryCatch(
  error = function(cnd) {
    base::paste0("~~", base::conditionMessage(c = cnd), "~~")
  },
  base::stop("This is an error", call. = FALSE)
)
```

In the code above, `base::stop("This is an error", call. = FALSE)` creates a condition object which is passed to the handler as its single argument, and the control is never returned to the code. This is different from `base::withCallingHandler()`, which returns the control to where the condition was signaled and `base::stop("This is an error", call. = FALSE)` is run:

```{r, eval=FALSE}
withCallingHandlers(
  error = function(cnd) {
    base::paste0("~~", base::conditionMessage(c = cnd), "~~")
  },
  base::stop("This is an error", call. = FALSE)
)
# - Error: This is an error
```

The function `tryCatch()` has one other argument `finally`, which is an expression to be evaluated before returning or exiting. This block of code runs regardless of whether the initial expression succeeds or fails:

```{r}
tryCatch(
  error = function(cnd) {
    base::paste0("~~", base::conditionMessage(c = cnd), "~~")
  },
  base::stop("This is an error", call. = FALSE),
  finally = {
    print("I run regardless")
  }
)
```

---

# Calling handlers

Exiting and calling handlers use "handlers" in different ways:

* Exiting handlers cause code to exit once the condition object has been created and passed to the handler function.

* Calling handlers return control to the context of where the condition object was created and the code continues normally. This makes calling handlers work well with non-error conditions.

The example below demonstrates the difference:

```{r}
# Exiting handler never returns control to the context where the condition is signaled
tryCatch(
  message = function(message) cat("Caught a message \n"),
  {
    base::message("one")
    base::message("two")
  }
)
# Calling handlers will return control to the context where the condition is signaled
withCallingHandlers(
  message = function(message) cat("Caught a message \n"),
  {
    base::message("one")
    base::message("two")
  }
)
```

The return value of a calling handler is ignored since the code continues to run after the handler completes. One unique side-effect of calling handlers is that we can **muffle** the condition signal. By default, a condition will propagate to the parent handlers at the top level, all the way up to the default handler: 

```{r}
# Propagation of condition
withCallingHandlers(
  # Default calling handler
  message = function(message) {
    cat("level 2\n")
  },
  withCallingHandlers(
    # Calling handler within the default handler
    message = function(message) {
      cat("level 1\n")
    },
    # This code creates the condition object to be passed on to level 1 calling handler
    # Then, after level 1 calling handler completes, the code is run and prints "Hello"
    # This creates another condition object to be passed on to level 2 default calling handler
    # Finally, the default calling handler completes and "Hello" is printed normally
    message("Hello")
  )
)
```

Another example of such propagation:

```{r}
# Propagation of condition
tryCatch(
  # Default calling handler
  message = function(message) {
    cat("level 2\n")
  },
  withCallingHandlers(
    # Calling handler within the default handler
    message = function(message) {
      cat("level 1\n")
    },
    # This code creates the condition object to be passed on to level 1 calling handler
    # Then, after level 1 calling handler completes, the code is run and prints "Hello"
    # This creates another condition object to be passed on to level 2 default calling handler
    # However, this time "Hello" is not printed since tryCatch() never returns control
    message("Hello")
  )
)
```

For the first example, we can muffle the default calling handler at level 2. In other words, the "Hello" message will not be printed:

```{r}
# Propagation of condition
withCallingHandlers(
  # Default calling handler
  message = function(message) {
    cat("level 2\n")
    # This prevents "message("Hello")" from creating a new condition object here
    rlang::cnd_muffle(message)
  },
  withCallingHandlers(
    # Calling handler within the default handler
    message = function(message) {
      cat("level 1\n")
    },
    # This code creates the condition object to be passed on to level 1 calling handler
    # Then, after level 1 calling handler completes, the code is run and prints "Hello"
    # This hello does not create a new condition object to be passed on to level 2 default
    # The default calling handler is muffled and the signal is not captured
    message("Hello")
  )
)
```

For the second example, we muffle the calling handler within the default calling handler:

```{r}
# Propagation of condition
withCallingHandlers(
  # Default calling handler
  message = function(message) {
    cat("level 2\n")
  },
  withCallingHandlers(
    # Calling handler within the default handler
    message = function(message) {
      cat("level 1\n")
      # This ensures control is not returned to where condition was signaled
      # The message "Hello" will not be printed
      # The default calling handler will not be triggered
      rlang::cnd_muffle(message)
    },
    # This code creates the condition object to be passed on to level 1 calling handler
    # However, after level 1 calling handler completes, the code is NOT run since its muffled
    message("Hello")
  )
)
```

---

# Call stacks

```{r}
# Sequence of function calls
f <- function() g()
g <- function() h()
h <- function() message("!")
# Calling f() returns the message
f()
```

Calling handlers are called in the context of the call that signaled the condition:

```{r}
withCallingHandlers(
  f(),
  message = function(cnd) {
    # Call stack tree
    lobstr::cst()
    # Do not create a conditional object
    cnd_muffle(cnd)
  }
)
```

Examine the tree above, the `function (cnd) lobstr::cst()` is linked to the context in which `f()` is called via the dashed line. On the other hand, exiting handlers are called in the context of the call to `tryCatch()`:

```{r}
tryCatch(
  f(),
  message = function(cnd) {
    lobstr::cst()
  }
)
```

```{r}
rlang::env_unbind(
  env = global_env(),
  nms = names(x = global_env())
)
```


---

# Exercises

## Exercise 1

What extra information does the condition generated by `abort()` contain compared to the condition generated by stop(), i.e. what’s the difference between these two objects? Read the help for ?abort to learn more.

```{r}
# The rlang condition object
rlang_cnd <- rlang::catch_cnd(
  expr = rlang::abort(message = "help")
)
str(rlang_cnd)
# Built-in condition object
built_in_cnd <- rlang::catch_cnd(
  expr = base::stop("help", call. = FALSE)
)
str(built_in_cnd)
```

**Similarities**

* Both objects are list objects. However, the built-in condition object has two elements compared to the `rlang` condition objects, which has three.

* Both objects have a "call" or "parent" element, which is the call that triggered the condition. Both are NULL since I suppressed the printing of the call in `stop()` and `rlang::abort()` by default does not print the call.

* Both objects have a "message" element.

**Differences**

* The `rlang` condition object has a "trace" element, which is in and of itself a list of three elements. This element is created by `rlang::trace_back()`.
  
  - A backtrack captures the sequence of calls that lead to the function that throws the error, this is a call stack. 
  
---

## Exercise 2

Predict the results of evaluating the following code:

```{r}
show_condition <- function(code) {
  tryCatch(
    error = function(cnd) "error",
    warning = function(cnd) "warning",
    message = function(cnd) "message",
    {
      code
      NULL
    }
  )
}
```

* The following will return "error", since `stop()` will create a condition object to be passed to the exiting handler matching with `error`. The Null value is not printed since the control is never passed back to the original code where the condition object is triggered:

```{r}
show_condition(stop("!"))
```

* The following will return the NULL object as no condition is signaled by the input `10L`. However, only the NULL object in {} is printed as 10L is not an expression but a vector:

```{r}
show_condition(10L)
```

* The following will print "warning" for the same reason as the first one:

```{r}
show_condition(warning("?!"))
```

* The following will print "message". The double vector 10 does not create a condition object but `message("?")` will and `warning("?!")` is never run since control is never returned back to this code once the message handler completes:

```{r}
show_condition({
  10
  message("?")
  warning("?!")
})
```

---

## Exercise 3

Explain the results of running this code:

```{r}
withCallingHandlers( # level 2
  message = function(cnd) message("b"),
  withCallingHandlers( # Level 1
    message = function(cnd) message("a"),
    {
      message("c")
    }
  )
)
```

* The code `message("c")` creates a condition object that is passed to level 1 calling handler, which prints `message("a")`. This, in turn, creates a condition object that is passed to level 2 calling handler at the toppest level, which prints `b` to the console instead of `message("b")` as there is no more calling handler on top to capture the message condition.

* After level 2 calling handler completes, control is passed back to the context in which the condition is signaled, and level 1 calling handler is run. This leads to `message("c")` being captured and `a` being printed to the console. The character vector `a` is printed instead of `message("a")` as it is no longer caught by level 1 calling handling since handlers are applied in order. Or else we would be stuck in an infinite loop.

* After level 1 calling handler completes, control is passed back to the context in which the condition is signaled, and `message("c")` is run. This triggers level 2 calling handler again and so `b` is printed to console.

* After level 2 calling handler completes another time, control is passed back to the context in which `message("c")` is called and `c` is finally printed to console since it is no longer captured as a condition by the calling handlers up top.

---

## Exercise 4

```{r}
catch_cnd
```

* The function `rep_named(names, x)` repeats a value along a names vectors:

```{r}
# One element character vector produces a list with one vector element
rep_named(names = c("condition"), x = list(identity))
```

---

The original source code of `catch_cnd()` was much simpler:

```{r, eval=FALSE}
catch_cnd <- function(expr) {
  tryCatch(
    condition = function(cnd) cnd,
    {
      force(expr)
      return(NULL)
    }
  )
}
```

---

## Exercise 5

Rewrite `show_condition()` to use a single handler:

```{r}
show_condition2 <- function(code, class) {
  if (class == "error") {
    tryCatch(
      error = function(cnd) "error",
      {
        code
        NULL
      }
    )
  } else if (class == "message") {
    tryCatch(
      message = function(cnd) "message",
      {
        code
        NULL
      }
    )
  } else if (class == "warning") {
    tryCatch(
      warning = function(cnd) "warning",
      {
        code
        NULL
      }
    )
  } else {
    rlang::abort(
      message = "Class must either by 'error,' 'message,' or 'warning.'"
    )
  }
}
```

Let us see it in action:

```{r}
show_condition2(code = stop("!"), class = "error")
```

Another example:

```{r}
# Since the class and condition are mismatched, the code runs normally
show_condition2(code = message("!"), class = "error")
```

Another example:

```{r}
show_condition2(code = message("!"), class = "message")
```

---

Another solution:

```{r}
show_condition3 <- function(code) {
  tryCatch(
    condition = function(cnd) {
      if (inherits(x = cnd, what = "error")) {
        return("error")
      }
      if (inherits(cnd, "warning")) {
        return("warning")
      }
      if (inherits(cnd, "message")) {
        return("message")
      }
    },
    {
      code
      NULL
    }
  )
}
```

In this implementation, `inherits(x = cnd, what = "class")` returns `TRUE` if `cnd` is indeed a condition object of the class "class". This is essentially checking if `class(cnd)` is "class", returning `TRUE` or `FALSE` depending on the outcome. This solution is shorter than my original solution.

---

Let us see it in action:

```{r}
show_condition3(code = stop("!"))
```

Another example:

```{r}
show_condition3(code = message("!"))
```
