---
title: "Signalling conditins"
author: "Yang Wu"
date: "9/12/2021"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.align = "center")
library(rlang)
library(tidyverse)
options(digits = 4)
```

# Errors

In base R, errors are thrown by `base::stop(..., call. = TRUE)`. It is good practice to suppress the part of the error message that includes the call to the function that generated the error, since we can usually get the information using `base::traceback()`.

```{r, eval=FALSE}
base::stop("This is an error", call. = FALSE)
```

The `rlang` equivalent is `rlang::abort()`, which automatically suppresses printing the call.

```{r, eval=FALSE}
rlang::abort(message = "This is an error")
```

To paste together multiple inputs as complex error messages, we could use `glue::glue(..., .sep = "", envir = parent.frame(), .open = "{", .close = "}", .na = "NA", .transformer = identity_transformer, .trim = TRUE)`, which formats and interpolate expressions enclosed by braces as R code.

```{r}
intro <- function(name, profession, country) {
  glue::glue("My name is {name}, a {profession} from {country}.")
}
intro("Ken", "Data Analyst", "China")
```

Examine the [tidyverse style guide](https://style.tidyverse.org/error-messages.html) for general principles for error messages.

---

# Warnings

Warnings can be created using `base::warning(..., call. = TRUE)`, but use `rlang::warn()` to suppress printing the call. The function below does not show the warnings since I set `knitr::opts_chunk$set(warning = FALSE)`: 

```{r}
warning_function <- function() {
  cat("1\n")
  base::warning("warning 1")
  cat("2\n")
  base::warning("warning 2")
  cat("3\n")
  base::warning("warning 3")
}
warning_function()
```

We can control the behavior of warnings through options:

* To make the warnings appear immediately, `options(warn = 1)`

* To turn warnings into errors, `options(warn = 2)`

* To restore to default, which means that warnings are cached and printed when the control returns to the top level interactive environment (usually the global env), `options(warn = 0)`

---

# Messages 

Messages are informational; they are used to inform the function user that the function has done something on their behalf.

```{r}
message_function <- function() {
  cat("1\n")
  base::message("message 1")
  cat("2\n")
  base::message("messsage 2")
  cat("3\n")
  base::message("message 3")
}
message_function()
```

The `rlang` equivalent is `rlang::inform()`.

---

# Exercises 

## Excercise 1

Write a wrapper around `file.remove()` that throws an error if the file to be deleted does not exist. We will use `file.exists`, which returns a logical vector indicating whether the files named by its argument exist.

```{r}
file_remove2 <- function(file) {

  # Stop of the file does not exist
  if (!file.exists(file)) {
    rlang::abort(message = glue::glue("The file {file} does not exist. There is nothing to delete."))
  }

  file.remove(file)
}
```

```{r, eval=FALSE}
# Test
file_remove2(file = "ken")
# Error: The file ken does not exist. There is nothing to delete.
```

---

## Exercise 2

What does the `appendLF` argument to `message()` do? How is it related to `cat()`?

**Answer**: The argument `appendLF` is a logical vector of length 1 indicating whether messages given as a character string should have a newline appended.

```{r}
multiline_msg <- function(appendLF = TRUE) {
  message("first", appendLF = appendLF)
  message("second", appendLF = appendLF)
  cat("third")
  cat("fourth")
}
```

Test:

```{r}
multiline_msg(appendLF = FALSE)
multiline_msg(appendLF = TRUE)
```

On the other hand, `cat` uses `\n` for a new line:

```{r}
cat("d", sep = "\n", "z")
```

---

# Ignoring conditions

Messages and warnings can be suppressed:

```{r}
# Warnings as expressions
{
  warning("Uhoh!")
  warning("Another warning")
  1
}
# Suppress warning
base::suppressWarnings(expr = {
  warning("Uhoh!")
  warning("Another warning")
  1
})
# Messages as expressions
{
  message("Hello there")
  2
}
# Suppress message
base::suppressMessages(expr = {
  message("Hello there")
  2
})
# Suppress warning does not suppress message
suppressWarnings({
  message("You can still see me")
  3
})
```

---

We can also ignore conditions by using `base::try()`, which tries an expression by allowing error to recover. A useful pattern is to do assignment inside the call to `try()`. This lets you define a default value to be used if the code does not succeed. This works because the argument `expr` in `try()` is evaluated in the calling environment, which in the case below is the global env, not inside the function.

```{r}
default <- NULL
try(expr = {default <- read.csv("possibly-bad-input.csv")}, silent = TRUE)
```


