---
title: "Recursing Over Environments"
author: "Ken Wu"
date: "8/24/2021"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.align = "center")
library(tidyverse)
library(rlang)
options(digits = 4)
```

# Recursing over environments {#recursing}

It is natural to work with environments recursively:

```{r, eval=FALSE}
f <- function(..., env = rlang::caller_env()) {
  if (identical(env, rlang::empty_env())) {
    # Base case or stop when this case is reached
  } else if (success) {
    # Success case
  } else {
    # Recursive case
    # Call this function recursively now with the parent environment of the current environment
    f(..., env = rlang::env_parent(env))
  }
}
```

---

This function can be written to find the environment `where()` a binding, or name is defined starting with the `rlang::caller_env()`:

```{r}
where <- function(name, env = rlang::caller_env()) {
  # If the env is the empty environment
  if (identical(env, rlang::empty_env())) {
    # Base case or stop case
    base::stop("Can't find ", name, call. = FALSE)
    # If the name/binding exists in this environment, return this environment
  } else if (rlang::env_has(env, name)) {
    # Success case
    env
  } else {
    # If "name" is not found in the current environment, call this function but now check the parent
    # Recursive case
    where(name, rlang::env_parent(env))
  }
}
```

* The function `base::stop(..., call. = TRUE, domain = NULL)` has the argument "call.", which is a logical indicating if the call should become part of the error message.

---

The function `where()` has three cases:

* The base case: we’ve reached the empty environment and haven’t found the binding. We can’t go any further, so we throw an error.

* The successful case: the name exists in this environment, so we return the environment.

* The recursive case: the name was not found in this environment, so try the parent.

```{r, eval=FALSE}
# This should throw an error since the name is not defined at all
where(name = "ken", env = rlang::caller_env())
```
```{r}
# This should return the global environment
x <- "ken"
where(name = "x", env = rlang::caller_env())
# This should return the ggplot2 package as the binding for ggplot() is in the package
where(name = "ggplot", env = rlang::caller_env())
```

---

Let us see another example that is a bit more complex: 

```{r}
# Define a new environment whose parent is the global environment
# This env has two bindings
e1 <- rlang::env(rlang::global_env(), x = 3, ken = "gets a job")
# Check that its parent is indeed the global env
rlang::env_parent(env = e1)
# Create another new env whose parent is e1 with two bindings
e2 <- rlang::env(e1, I = "need a job", x = sample(x = 1:23, size = 10, replace = TRUE))
# Address of e1
e1
# Check that e2's parent is indeed e1
rlang::env_parent(env = e2)
```

Now we will use `where()` to search for bindings:

```{r}
# The addresses of e1 and e2
e1
e2
# Search for the object ken starting with e2
where(name = "ken", env = e2)
```

Note that `where(name = "ken", env = e2)` does not find `ken` in `e2` and so it looks in its parent, `e1`, and finds it there.

```{r}
# Create a binding in the global env
job <- "can't be found"
# Search for object job starting with e2
where(name = "job", env = e2)
```

Here, `where(name = "job", env = e2)` does not find `job` in `e2` and so it looks in its parent, `e1`. It does not find `job` in `e1` and so it scopes for it in the parent, which is the global environment and finds it there. If we try to scope for a binding that does not exist at all, (including in all the packages currently loaded), then the function will throw an error.

---

# Iteratioin versus recursion

It is possible to re-write the recursive function from section \@ref(recursing) as a while loop:

```{r, eval=FALSE}
f2 <- function(..., env = rlang::caller_env()) {
  # If identical() evaluates to false, i.e., not the empty env, then execute the following
  # The ! negates false to true to ensure the while loop executes
  while (!identical(x = env, rlang::empty_env())) {
    # Now evaluate if the case is a success, and if so, return the env
    if (success) {
      base::return()
    }
    # If not the success case, then modify to search the parent
    env <- env_parent(env)
  }

  # Base base or stop case
  # The is reached when the while condition evaluates to false, i.e., is the empty env,
  # Still no success case so throw an error
  rlang::stop(..., call. = FALSE)
}
```

---

# Exercises 

## Exercise 1

Modify `where()` to return all environments that contain a binding for name. Carefully think through what type of object the function will need to return:

```{r}
# The pre-allocated list container for all env's where "name" is found is initialized via default arg
where2 <- function(name, env = rlang::caller_env(), container = vector(mode = "list")) {

  # If the env is the empty environment
  if (identical(env, rlang::empty_env())) {

    # Just return the empty list since no env contains the binding "name"
    container
    # If the name/binding exists in this environment, store this env in the list container
  } else if (rlang::env_has(env, name)) {

    # The use of c() combines several lists into one list (see chapter 3 on list/atomic vectors)
    container <- c(container, env)
    # Continue searching its ancestors
    # Note we need to now specify the updated container by overriding the default arg
    where2(name, rlang::env_parent(env), container = container)
  } else {

    # If "name" is not found in the current environment, call this function but now check the parent
    # Again, override the default arg using the same empty container
    # We do not wish to re-initialize the empty container each time the function is called
    where2(name, rlang::env_parent(env), container = container)
  }
}
```

Let us see it in action:

```{r}
# The binding "x" exists in all three env's--- e1, e2, and global
rlang::env_names(e1)
rlang::env_names(e2)
rlang::env_names(rlang::global_env())
# Now the function where should return a single list containing e1, e2, and global
where2(name = "x", env = e2)
# Let us check the addresses
e1
e2
```

---

This is the textbook's solution:

```{r,eval=FALSE}
where2 <- function(name, env = caller_env(), results = list()) {
  if (identical(env, empty_env())) {
    # Base case
    results
  } else {
    # Recursive case
    if (env_has(env, name)) {
      results <- c(results, env)
    }
    where2(name, env_parent(env), results)
  }
}
```


---

## Exercise 2

Write a function called `fget()` that finds only function objects. It should have two arguments, `name` and `env`, and should obey the regular scoping rules for functions: if there’s an object with a matching name that’s not a function, look in the parent. For an added challenge, also add an `inherits` argument which controls whether the function recurses up the parents or only looks in one environment:

```{r}
fget <- function(name, env = rlang::caller_env(), inherits) {

  # If we do not need to check parents
  if (inherits == FALSE) {
    # Check if the empty env is reached
    if (identical(x = env, y = rlang::empty_env())) {
      stop("Cannot find function object", call. = FALSE)
      # If binding exists AND is a function, extract it
    } else if (rlang::env_has(env, nms = name)) {
      obj <- rlang::env_get(env = env, nm = name)

      # If the object is a function, return it, if not, do nothing
      if (rlang::is_function(obj)) {
        return(obj)
      }
      # If not found in current env, stop and throw an error
    } else {
      stop("Cannot find function object", call. = FALSE)
    }
  }

  if (inherits == TRUE) {
    # Check if the empty env is reached
    if (identical(x = env, y = rlang::empty_env())) {
      stop("Cannot find function object", call. = FALSE)
      # If binding exists AND is a function, extract it
    } else if (rlang::env_has(env, nms = name)) {
      obj <- rlang::env_get(env = env, nm = name)

      # If the object is a function, return it, if not, do nothing
      if (rlang::is_function(obj)) {
        return(obj)
      }
      # If not found, check parent
    } else {
      fget(name, env = rlang::env_parent(env), inherits = inherits)
    }
  }
}
```

Let us see it in action:

```{r}
fget(name = "ggplot", env = e2, inherits = TRUE)
fget(name = "sum", env = e2, inherits = TRUE)
fget(name = "where", env = rlang::global_env(), inherits = FALSE)
```

---

Here is the textbook's solution:

```{r}
fget_text <- function(name, env = caller_env(), inherits = TRUE) {

  # If the binding exists, store it as an object
  if (env_has(env, name)) {
    obj <- env_get(env, name)

    # If the object is a function object, then return it
    if (is.function(obj)) {
      return(obj)
    }
  }

  # If empty env is reached OR inherits == FALSE, throw an error that no function is found
  if (identical(env, emptyenv()) || !inherits) {
    stop("Could not find a function called \"", name, "\".",
      call. = FALSE
    )
  }

  # If env_has(env, name) evaluate to FALSE and (identical(env, emptyenv()) || !inherits) == FALSE
  # Keep searching for the function object thru parents
  fget_text(name, env_parent(env))
}
```

Let us see this function in action:

```{r}
fget_text(name = "ggplot", env = e2, inherits = TRUE)
fget_text(name = "sum", env = e2, inherits = TRUE)
```

---

```{r}
# Unbind all in global
rlang::env_unbind(env = rlang::global_env(), nms = base::names(rlang::global_env()))
```
