---
title: "Subsetting"
author: "Ken Wu"
date: "6/20/2021"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(collapse = TRUE)
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(fig.width = 5, fig.height = 3, fig.align = "center")
library(tidyverse)
```

## Selecting multiple elements

### Atomic Vectors

There are six ways to subset atomic vectors:

```{r}
# Number after decimal represents original position in the vector
x <- c(2.1, 4.2, 3.3, 5.4)
```


* **Positive integers** returns the elements at the specified positions in the vector, displaying the elements in the order in which they appear in the subsetting operator "[":

```{r}
# Reference
x
# Select fourth, second, and first elements, in that order
x[c(4, 2, 1)]
```

The order() function returns a permutation which rearranges its first argument into ascending or descending order, breaking ties by further arguments; sort.list() does the same, using only one argument. 

```{r}
# Reference
x
# Generate order vector descending
order(x, decreasing = TRUE)
# Select fourth, second, third, and first elements, in that order
x[order(x, decreasing = TRUE)]
```

Duplicate indices will duplicate values and real numbers as indices are silently truncated to integers:

```{r}
# Duplicate
x[c(2, 2, 4, 4)]
# Real numbers
x[c(2.9, 2.2, 4.8, 4.7, 3.5)]
# These doubles are coerced to integers before indexing
as.integer(c(2.9, 2.2, 4.8, 4.7, 3.5))
```

* **Negative integers** exclude elements at the specified positions, leaving behind the unspecified elements in their original positions. We cannot mix positive and negative integers in a single subset:

```{r}
# Reference
x
# Exclude the fourth and second elements
x[-c(1, 4)]
```

* **Logical vectors** select elements where the corresponding logical value is TRUE. We can select based on conditions.

```{r}
# Select the first and fourth elements
x[c(TRUE, FALSE, FALSE, TRUE)]
# Select randomly from x
x[as.logical(x = sample(x = c(0, 1), size = 4, replace = TRUE))]
# Select elements that are greater than 2
x[x > 2]
# Select first twice
# Combining a double and logical values results in a double vector
x[c(1, TRUE, 3, FALSE)]
```

In x[y], when x and y are different lengths, the recycling rules is applied; the shorter of the two is recycled to the length of the longer. 

```{r}
# Select first and third, since the TRUE FALSE combo is recycled
x[c(TRUE, FALSE)]
# Equivalent to
x[c(TRUE, FALSE, TRUE, FALSE)]
```

An NA missing values in the index always returns a missing value in the output:

```{r}
x[c(TRUE, FALSE, NA, NA)]
```

* **Nothing** returns the original vector.

```{r}
x[]
```

* **Zero** returns a zero-length vector.

```{r}
# Zero
x[0]
# NULL
x[NULL]
```

* If an atomic vector is named, we can select using **character vectors** to return elements with matching names (when subsetting with [, names are always matched exactly):

```{r}
# Create a new atomic vector
y <- runif(n = 10, 5, 8)
# Name
y <- setNames(object = y, letters[3:12])
# Attribute
attributes(y)
# Structure
str(y)
# Type
typeof(y)
# Reference
y
# Select f through j
y[c("f", "g", "h", "i", "j")]
# Like integer indices, you can repeat indices
y[c("g", "g", "i")]
```

Factors are not treated specially, and so subsetting will use the underlying integer vector, not the character levels.

```{r}
rm(list = ls())
```


## Lists

Lists work in the same way as subsetting an atomic vectors, using [ always returns a list.

---

## Matrices and arrays

Three ways to subset higher-dimensional structures:

* With multiple vectors
* With a single vector
* With a matrix

1. The most common way of subsetting matrices and arrays is to supply a 1D index for each dimension, separated by a comma. Blank subsetting allows for keeping all rows, or all columns.

```{r}
# Create a 3 x 3 matrix
matrix <- matrix(data = 1:9, nrow = 3)
# Name the columns
colnames(matrix) <- c("A", "B", "C")
# Reference
matrix
# Select the first row of first two columns (Using multiple vectors)
# One 1D index for each dimension (returns an integer vector)
matrix[1, c(1, 2)]
# Type
typeof(matrix[1, c(1, 2)])
```

* Using positive integer vectors

```{r}
# Select all rows of column two (returns an integer vector)
matrix[, 2]
# Type
typeof(matrix[, 2])
```

* Using negative integer vectors

```{r}
# Exclude first and third rows and keep all columns
matrix[-c(1, 3), ]
# Exclude third row and third column
matrix[-3, -3]
# Returns a matrix
is.matrix(matrix[-3, -3])
```


* Using logical vectors and character vectors

```{r}
# Select first two rows from first and third columns
matrix[c(TRUE, TRUE, FALSE), c("A", "C")]
# Returns a matrix
is.matrix(matrix[c(TRUE, TRUE, FALSE), c("A", "C")])
```

By default, matrix subsetting using [ simplifies the results to the lowest possible dimensionality. 

2. Since matrices and arrays are vectors with special attributes, we can subset using a single vector as if they were a 1D vector. The outer() function computes the outer product of the arrays X and Y. The outer product of the arrays X and Y is the array A with dimension c(dim(X), dim(Y))

```{r}
# Create 5 x 5 Matrix
vals <- outer(X = 1:5, Y = 1:5, FUN = "paste", sep = ",")
# Dimension
dim(vals)
# Structure
str(vals)
# Attributes
attributes(vals)
# Reference
vals
```

The matrix "vals" is essentially a 25-element vector with dim attributes. So we can subset using a single vector:

```{r}
# Reference
as.vector(x = vals)
# Length
length(as.vector(x = vals))
# Select the fourth element and the 24th element (returns a character vector)
vals[c(4, 24)]
is.character(vals[c(4, 24)])
# Exclude elements 3, 4, and 10
vals[-c(3, 4, 10)]
is.character(vals[-c(3, 4, 10)])
```

3. Higher dimensional data structures can be subsetted by an integer matrix or character matrix. Each row in the index matrix specifies the location of one value (e.g. the i-jth element of an n x m matrix) and each column in the index matrix corresponds to a dimension in the array. That is, the number of columns in the index matrix must be =< the one less the number of columns in the higher dimensional data structures that needs to be subsetted. In our case, "vals" is a 3 x 3 matrix and so the index matrix must have $=< (3 - 1) = 2$ columns.

```{r}
# Create index matrix
set.seed(5)
index_matrix <- matrix(
  data = as.integer(runif(n = 6, min = 1, max = 5)),
  ncol = 2, byrow = TRUE
)
# Select the following elements
# Row 1 and column 3
# Row 4 and column 2
# Row 1 and column 3
index_matrix
# Reference
vals
# Subset
vals[index_matrix]
```

---

## Data frames

* When subsetting with a single index, data frames behave like lists and index the columns; so df[1:2] selects the first two columns.

* When subsetting with two indices, data frames behave like matrices, so df[1:3, ] selects the first three rows and all columns. R recognizes dimensions in terms of rows and columns; that is, n x m means n rows and m columns.

```{r}
# Create data frame
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])
# Row names
row.names(df)
# Show data frame
df
```

1. Select rows using logical expression and integer vectors:

```{r}
# Select all rows where the value in column "x" is 2
df[df$x == 2, ]
# Select first and third rows and all columns
df[c(1, 3), ]
```

2. Select columns like a list:

```{r}
# Select columns using character vector, in that order (single index)
df[c("y", "x")]
# Returns a list
typeof(df[c("y", "x")])
# Preserves data frame
is.data.frame(df[c("y", "x")])
# Exclude columns one and two
df[-c(1, 2)]
```

3. Select columns like a matrix:

```{r}
# Select columns "y" and "x", in that order (double index)
df[, c("y", "x")]
# Returns a list
typeof(df[, c("y", "x")])
# Preserves data frame
is.data.frame(df[, c("y", "x")])
# Exclude columns "z and "y'
df[, -c(2, 3)]
# Does not preserve data frame
is.data.frame(df[, -c(2, 3)])
```

There's an important difference if you select a single column: matrix subsetting simplifies by default, list subsetting does not.

```{r}
# List subsetting preserves data frame structure
str(df["z"])
# Matrix subsetting does not preserve data frame structure
str(df[, "z"])
```

---

## Tibbles

Subsetting a tibble with [ always returns a tibble:

```{r}
# Create a tibble
df <- tibble(x = 1:3, y = 3:1, z = letters[1:3])
# List subsetting
str(df["z"])
# Matrix subsetting
str(df[, "z"])
```

---

### Preserving dimensionality

By default, subsetting a matrix or data frame with a single number, a single name, or a logical vector containing a single TRUE, will simplify the returned output, i.e. it will return an object with lower dimensionality. 

1. For matrices and arrays, any dimensions with length 1 will be dropped:

```{r}
# Select a single element in the matrix returns a 1D integer vector of length 1
matrix[2, 3]
# Structure
is.integer(matrix[2, 3])
# Length
length(matrix[2, 3])
# Dimensions (1 x 1 scalar so both dims get dropped)
dim(matrix[2, 3])
# Preserve dimensionality
matrix[2, 3, drop = FALSE]
# Now the dimensions are preserved
dim(matrix[2, 3, drop = FALSE])
```

* Example 2:

```{r}
# Select columns 1 through 3 returns an 1D integer vector
matrix[, 3]
# Structure
str(matrix[, 3])
# Preserve dimensionality
matrix[, 3, drop = FALSE]
# Now the dimensions are preserved
dim(matrix[, 3, drop = FALSE])
```

2. Data frames with a single column will return just that column:

```{r}
# Create data frame
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])
# Select column x using double indices
df[, "x"]
# Structure
str(df[, "x"])
# Preserve dimensionality
str(df[, "x", drop = FALSE])
```

---

## Exercises

### Exercise 1

```{r}
# Select all rows where the cyl column equals 4 (i.e. all 4 cylinder cars)
mtcars[mtcars$cyl == 4, ]
# Exclude columns 1 through 4
mtcars[-c(1:4)]
# Select all rows where the cyl column is <= 5
mtcars[mtcars$cyl <= 5, ]
# Select all rows where the cyl column equals 4 or 6
mtcars[mtcars$cyl == 4 | mtcars$cyl == 6, ]
# Another way is to use %in% like the IN command in SQL
mtcars[mtcars$cyl %in% c(4, 6), ]
```

---

### Exercise 2

```{r}
# Create integer vector
x <- 1:5
# Select NA
x[NA]
# Select NA_integer type
x[NA_integer_]
```

In contrast to NA_real_ or NA_integer_, NA has logical type and logical vectors are recycled to the same length as the numeric vector x that is being subset, i.e. x[NA] is recycled to x[NA, NA, NA, NA, NA].

---

### Exercise 3

The function upper.tri() returns a matrix of logical values the same size of a given matrix with entries TRUE in the lower or upper triangle.

```{r}
# Create a matrix
x <- outer(X = 1:5, Y = 1:5, FUN = "*")
x
# Return a logical matrix
upper.tri(x)
# Return entries that are TRUE in the index (logical) matrix
# That is, those that are above the diagonal
x[upper.tri(x)]
```

When subsetting with logical matrices, all elements that correspond to TRUE will be selected. Matrices extend vectors with a dimension attribute, so the vector forms of subsetting can be used (including logical subsetting). We should take care, that the dimensions of the subsetting matrix match the object of interest — otherwise unintended selections due to vector recycling may occur. Please also note, that this form of subsetting returns a vector instead of a matrix, as the subsetting alters the dimensions of the object.

---

### Exercise 4

The line mtcars[1:20] returns an error since the single index indicates the columns of the data frame, which behaves like a list. There are less than columns in the mtcars data set, and so the indexing is out-of-bounds. On the other hand,  the line mtcars[1:20, ] selects the first 20 rows and all columns from the data set. In this case the data frame is subsetted with two vectors, and so it behaves like a matrix.

---

### Exercise 5

The function:
```{r}
# A function for extracting diagonal entries of a matrix
extract_diag <- function(x) {

  # Number of diagonal entries
  n <- min(nrow(x), ncol(x))

  # Index matrix
  # Each row represents the i-jth entry
  # In this case, it is (1,1), (2,2), ...
  indices <- cbind(seq_len(n), seq_len(n))

  # Extract corresponding entries
  x[indices]
}
```

Test:

```{r}
# Create a 4 x 4 matrix
x <- matrix(data = runif(n = 14, min = 1, max = 202), ncol = 4)
x
# Extract diagonal entries
extract_diag(x)
```

---

### Exercise 6

```{r}
# Create data frame
df <- data.frame(x = 1:4, y = c(2, NA, 4, 5), z = c("a", "b", "c", NA))
df
# Return missing values
is.na(df)
# Select missing values
df[is.na(df)]
# Replace NA with 0
df[is.na(df)] <- 0
df
```

This expression replaces the NAs in df with 0. Here is.na(df) returns a logical matrix that encodes the position of the missing values in df. Subsetting and assignment are then combined to replace only the missing values.
