---
title: "Selecting a single element"
author: "Ken Wu"
date: "6/21/2021"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(collapse = TRUE)
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(fig.width = 5, fig.height = 3, fig.align = "center")
library(tidyverse)
```

## Two other subsetting operators [[ and $

[[ is most important when working with lists because subsetting a list with [ always returns a smaller list:

```{r}
# Create a list
x <- list(1:3, "a", 4:6)
str(x)
# Subsetting using [ returns a smaller list
str(x[3])
# Subsetting using [[ returns the object (a vector) the list element points to
str(x[[3]])
```

When extracting multiple (or even zero!) elements, you have to make a smaller list by subletting with the [ operator. The [[ operator must be used with a single positive integer or a single string. If a vector is used with [[ then it will subset recursively:

```{r}
x[[c(1, 3)]]
# Equivalent to
x[[1]][[3]]
```

For consistency, use [[ to extract a single value with atomic vectors too. Use [ to extract multiple elements.

---

The "$" operator is a shorthand operator:

```{r}
mtcars$mpg
# Equivalent to
mtcars[["mpg"]]
# Type
typeof(mtcars[["mpg"]])
```

One difference between $ and [[ is that the former uses left-to-right partial matching:

```{r}
# Create a list
x <- list(obc = 1)
# Partial matching
x$o
# No partial matching
x[["o"]]
```

---

## purrr::pluck() and purrr::chuck()

```{r}
# Create a named list
x <- list(
  a = list(1, 2, 3),
  b = list(3, 4, 5)
)
# Structure
str(x)
# Attributes
attributes(x)
# Names is a character vector (metadata)
is.character(names(x))
```

When the element is missing, pluck() always returns NULL (or the value of the .default argument) and chuck() always throws an error. The function pluck() also allows you to mix integer and character indices, and provides an alternative default value if an item does not exist:

* Example 1

```{r}
# Select the object list "a" from the list x
# From list "a", select the first element
purrr::pluck(x, "a", 1)
# Structure
str(purrr::pluck(x, "a", 1))
# Returns a double vector of length 1
is.double(purrr::pluck(x, "a", 1))
# Using chuck()
purrr::chuck(x, "b", 2)
```

* Example 2

```{r}
# Missing element returns NULL
purrr::pluck(x, "c", 1)
# Alternative default value 
purrr::pluck(x, "c", 1, .default = "Does not exist")
```

---

## Exercises

### Exercises 1 

```{r}
# Data set
as.tibble(mtcars)
# Select third value from the cyl variable
mtcars[["cyl"]][[3]]
# Second 
mtcars[[2]][[3]]
# Third 
mtcars[[c(2, 3)]]
# Fourth 
mtcars$cyl[[3]]
# Fifth 
purrr::pluck(mtcars, "cyl", 3)
# Sixth 
purrr::chuck(mtcars, "cyl", 3)
# Seventh 
mtcars[2]$cyl[[3]]
# Eighth
mtcars[, 2][[3]]
# Select the "cyl" column and the third value from the vector
mtcars[, "cyl"][[3]]
# Tenth (View tibble as a matrix and subset using two indices)
mtcars[3, 2]
# Select third row and select the second column
mtcars[3, ][, 2]
# Twelveth
with(mtcars, cyl[[3]])
```

---

### Exercise 2

```{r}
# Model
model <- lm(mpg ~ wt, data = mtcars)
# Type
typeof(model)
# Attributes
attributes(model)
# Extract DF residual (A single value)
model[["df.residual"]]
# Or alternatively
model$df.residual
# Type
typeof(model[["df.residual"]])
```

Second task:

```{r}
# Type of object that summary() returns
typeof(summary(model))
# Attributes
attributes(summary(model))
# Extract R squared (Another )
summary(model)[["r.squared"]]
# Or alternatively
summary(model)$r.squared
# Type
typeof(summary(model)[["r.squared"]])
```

---

## Subsetting and assignment

The basic form is x[i] <- value:

```{r}
# Create a vector
x <- 1:5
# Select the fifth and third values, in that order,
# Replace them with 20 and 77, respectively
x[c(5, 3)] <- c(20, 77)
x
```

Ensure that length(value) is the same as length(x[i]) to avoid unexpected results. For lists, the syntax is x[[i]] <- NULL to remove a component. Use x[[i]] <- list(NULL) to add a literal NULL.

```{r}
# Create a list
x <- list(a = c(3, 4), b = 4, c = 7)
# Remove the first object
x[["a"]] <- NULL
# Add literal NULL
x[["b"]] <- list(NULL)
```

Subsetting with nothing can be useful with assignment because it preserves the structure of the original object. Compare the following two expressions. In the first, mtcars remains a data frame because you are only changing the contents of mtcars, not mtcars itself. In the second, mtcars becomes a list because you are changing the object it is bound to.

```{r}
# Subsetting with nothing
mtcars[] <- lapply(mtcars, as.integer)
is.data.frame(mtcars)
# No subsetting
mtcars <- lapply(mtcars, as.integer)
is.data.frame(mtcars)
```

```{r}
rm(list = ls())
```



